Supplemental Materials for the paper: \\Pesticides pollution of small streams in Germany

### 2016-09-14 article.tex
finalised precipitation and size width relationships

%% -------------------------------------------------------------------------
\section{Introduction}
More than 50\% of the total land area in Germany are used by agriculture \citep{statistisches_bundesamt_bodenflache_2014}.
In the year 2014 more than 45,000 tonnes of 766 authorized pesticides were sold for application on this area \citep{bundesamt_fur_verbraucherschutz_und_lebensmittelsicherheit_absatz_2015}.
The applied pesticides may enter surface waters via spray-drift, edge-off-field run-off or drainage \citep{stehle_probabilistic_2013,schulz_comparison_2001,liess_determination_1999}.
Especially run-off after heavy precipitation events has been shown to be one of the major input routes for pesticides \citep{schulz_field_2004}.
Once entered the surface waters they may have adverse effects on biota and ecosystem functioning \citep{schafer_thresholds_2012}. 
Although, it is known that pesticides pollution and its effect increase with agricultural land-use \todo{citation needed}, studies investigating the shape of this relationships are missing.

@@

Small waters (SW) comprise a major fraction of streams \citep{nadeau_hydrological_2007}, accommodate a higher proportion of biodiversity compared to larger freshwater systems \citep{davies_comparison_2008, biggs_report_2014} and play an important role in recolonization of disturbed downstream reaches \citep{liess_analyzing_2005, orlinskiy_forested_2015}.
However, SWB might be also at high risk of pesticide contamination from adjacent agricultural areas and lower dilution potential \citep{schulz_field_2004}.
It has been shown that SW are more polluted than bigger streams \citep{stehle_pesticide_2015,schulz_field_2004}.
Despite their relevance only a small fraction of studies were conducted on pesticide pollution of SW \todo{citation Lorenz et al.}, with only few large scale studies.

@@

\subsection{Characterization of chemical pollution}
We characterized pesticide pollution using regulatory acceptable concentrations (RAC) \citep{brock_linking_2010}.
RACs are derived during pesticide authorization and no unacceptable ecological effect are expected if the environmental concentration remains below this concentration.
The German Federal Environmental Agency provided RACs for the 105 compounds with highest detection rates (Supplement, Table S2). 
We expressed RACs as Risk Quotient (RQ):

@@

While agricultural land use and catchment size vary only between sites, this is not the case for precipitation that changes also with time.
Therefore, we modeled the effects of precipitation in a separate model.
RQ showed a skewed distribution with an excess of zeros (no pesticides with RAC detected). 
Therefore, we modeled these as two processes (one generating values below the limit of quantification (LOQ) and one generating values above LOQ) using a Zero-Adjusted Gamma (ZAGA) distribution \todo{Cite rigby 2010}:

@@

Changes in $\mu$ can be interpreted as changes in the absolute value of RQ, whereas changes in $\pi$ can be interpreted as changes in the probability that there is any risk.
We fitted this model separately to each compound with a RAC, with at least 1000 samples and more then 5\% of values above quantification limit (see Supplement, Table S3 for a list of compounds). 
We implemented the model using the gamlss packages \cite{}

\todo{Add definition of swb.}

@@

\subsection{Effect of precipitation on exposure}

The spatio-temporal intersection revealed that 5\% of the samples were taken at or after days with rainfall events greater than 10mm / day (Supplement, Figure S3).
Precipitation at the day before sampling increases $\mu$ and also $\pi$ for nearly all modeled compounds. 
The effects were less pronounced ($\pi$) or not directed ($\mu$) for precipitation at the day of sampling (Figure \ref{fig:fig5}, top).

Quarters 2-4 showed higher RQ than in Q1, except for the compounds  (Figure \ref{fig:fig5}, bottom)

\begin{figure}[ht]
  \includegraphics[width=0.95\textwidth]{figure5.pdf}
  \caption{Estimated coefficients and their 95\% CI for the model describe in equations \ref{eqn:eqn3} and \ref{eqn:eqn4} for each of the 24 analysed compounds (see Supplement Table . Left column: Effect on mean RQ; Right column: effect on the probability that RQ \textgreater 0.
  Coefficients where the CI encompasses zero are shown in grey.
  }
  \label{fig:fig5}
\end{figure}

\subsection{Pesticide pollution of small water bodies}

>>>>

%% -------------------------------------------------------------------------
\section{Introduction}
\todo{Check all references to supplement}.
More than 50\% of the total land area in Germany are used by agriculture \citep{statistisches_bundesamt_bodenflache_2014}.
In the year 2014 more than 45,000 tonnes of 766 authorized pesticides were sold for application on this area \citep{bundesamt_fur_verbraucherschutz_und_lebensmittelsicherheit_absatz_2015}.
The applied pesticides may enter surface waters via spray-drift, edge-off-field run-off or drainage \citep{stehle_probabilistic_2013,schulz_comparison_2001,liess_determination_1999}.
Especially run-off after heavy precipitation events has been shown to be one of the major input routes for pesticides \citep{schulz_field_2004}.
Once entered the surface waters they may have adverse effects on biota and ecosystem functioning \citep{schafer_thresholds_2012}. 
Although, it is known that pesticides pollution and its effect increase with agricultural land-use \todo{citation needed}, studies investigating the shape of this relationships are missing.

@@

Small waters (SW) comprise a major fraction of streams \citep{nadeau_hydrological_2007}, accommodate a higher proportion of biodiversity compared to larger freshwater systems \citep{davies_comparison_2008, biggs_report_2014} and play an important role in recolonization of disturbed downstream reaches \citep{liess_analyzing_2005, orlinskiy_forested_2015}.
However, SWB might be also at high risk of pesticide contamination from adjacent agricultural areas and lower dilution potential \citep{schulz_field_2004}.
It has been shown that SW are more polluted than bigger streams \citep{stehle_pesticide_2015,schulz_field_2004}.
Despite their relevance only a small fraction of studies were conducted on pesticide pollution of SW, with only few large scale studies \citep{lorenz_specifics_2016}.

@@

A clear definition of SWB in terms of catchment or stream size is currently lacking \citep{lorenz_specifics_2016}. 
The WFD defines SWB with a catchment size between 10 and 100 km\textsuperscript{2}, without further consideration of streams \textless 10km\textsuperscript{2}. 
\citet{lorenz_specifics_2016} defines SWB with catchment size \textless 10km\textsuperscript{2} as SWB.
Because of data scarcity of streams \textless 10~km\textsuperscript{2} (Figure \ref{fig:fig3}) we define in this study all streams below 25~km\textsuperscript{2} as SWB. This catchment size corresponds to a stream width of approximately 2~meters (see Supplement Figure S2).

\subsection{Characterization of pesticide pollution}
We characterized pesticide pollution using regulatory acceptable concentrations (RAC) \citep{brock_linking_2010}.
RACs are derived during pesticide authorization and no unacceptable ecological effect are expected if the environmental concentration remains below this concentration.
The German Federal Environmental Agency provided RACs for the 105 compounds with highest detection rates (Supplement, Table S2). 
We expressed RACs as Risk Quotient (RQ):

@@

While agricultural land use and catchment size vary only between sites, this is not the case for precipitation that changes also with time.
Therefore, we modeled the effects of precipitation in a separate model.
RQ showed a skewed distribution with an excess of zeros (no pesticides with RAC detected). 
Therefore, we modeled these as two processes (one generating values below the limit of quantification (LOQ) and one generating values above LOQ) using a Zero-Adjusted Gamma (ZAGA) distribution \cite{rigby_generalized_2005,stasinopoulos_gamlss.dist:_2016}:

@@

Changes in $\mu$ can be interpreted as changes in the absolute value of RQ, whereas changes in $\pi$ can be interpreted as changes in the probability that there is any risk.
We fitted this model separately to each compound with a RAC, with at least 1000 samples and more then 5\% of values above quantification limit (see Supplement, Table S3 for a list of compounds). 
We implemented the model using the gamlss package \cite{stasinopoulos_generalized_2007}.

@@

\subsection{Effect of precipitation on pesticide exposure}

The spatio-temporal intersection revealed that 5\% of the samples were taken at or after days with rainfall events greater than 10mm / day (Supplement, Figure S3).
Precipitation at the day before ($precip_{-1}$) sampling increased $\mu$ and also $\pi$ for nearly all modeled compounds. 
The effects were less pronounced ($\pi$) or not directed ($\mu$) for precipitation at the day of sampling ($precip_{0}$)(Figure \ref{fig:fig5}, top).

Quarters 2-4 showed generally higher $\mu$ than in Q1. 
Exceptions were Diflufenican, that showed lower RQ in Q2-Q4 compared to Q1.
Smaller negative effects were found for Clothianidin (Q2) and Isoproturon (Q3) (Figure \ref{fig:fig5}, bottom and Supplement Table S4).
We found similar effects on $\pi$.

\begin{figure}[h]
  \includegraphics[width=0.95\textwidth]{figure5.pdf}
  \caption{Estimated coefficients and their 95\% CI for the model described in equations \ref{eqn:eqn3} and \ref{eqn:eqn4} for each of the 24 analysed compounds (see Supplement Tables S3 and S4 for a list of compounds from top to bottom and their respective coefficients). Left column: Effect on mean RQ ($\mu$); Right column: effect on the probability that RQ \textgreater 0 ($\pi$).
  Coefficients where the CI encompasses zero are shown in gray colour.
  }
  \label{fig:fig5}
\end{figure}

\subsection{Pesticides in small water bodies}

### 2016-09-14 supplement/Supplemental_Materials.tex
fixed model description
finalised precipitation and size width relationships

@@

@@

\begin{figure}[h]
	\centering
	\includegraphics[width = 0.6\textwidth]{precip}
	\caption{Distribution of precipitation at sampling occasions. top: at sampling date. bottom: at day before sampling.}
	\label{fig:precip}
\end{figure}

\chapter{Are small agricultural waters more polluted compared to bigger streams?}

\section{RAC Exceedances}
\begin{figure}[h]
	\centering
	\includegraphics[width = 0.7\textwidth]{ezgagrirac}
	\caption{Raw data used for the model in equation xxx and Figure xxx of the main article. Color codes the number of RAC exceedances (on a log-scale). Grey points denote sites without any exceedance.}
	\label{fig:ezgagrirac}
\end{figure}
\todo{update refs}

>>>>

\chapter{Catchment size - stream width relationships}
We studied the relationship between catchment size based on three datasets containing this informations:
Data delivered by the federal state Thuringia, \citet{vos_organic_2015} and \citet{fernandez_effects_2015} (both from Rhineland-Palatinate).
We fitted to each dataset separately and to the combined dataset a power-function.
The resulting models are shown in Figure \ref{fig:size_width}.

\begin{figure}[h]
	\centering
	\includegraphics[width = 0.8\textwidth]{width_size}
	\caption[Relationship between catchment size and stream width.]{Relationship between catchment size and stream width. A power function has been fitted to each dataset separately and the combined dataset (black line and equation). The gray rectangle marks the estimated with for a catchment size of 25km\textsuperscript{2}.}
	\label{fig:size_width}
\end{figure}

@@

\begin{figure}[h]
	\centering
	\includegraphics[width = \textwidth]{temporal}
	\caption{Number of sampling occasions per year and month.}
	\label{fig:temporal}
\end{figure}

@@

\chapter{Effect of precipitation and season on RQ}
\begin{figure}[h]
	\centering
	\includegraphics[width = 0.6\textwidth]{precip}
	\caption{Distribution of precipitation at sampling occasions. top: at sampling date. bottom: at day before sampling.}
	\label{fig:precip}
\end{figure}

\input{keeptab}
\input{keeptab2}

\chapter{Pollution of small streams}
\begin{figure}[h]
	\centering
	\includegraphics[width = 0.7\textwidth]{ezgagrirac}
	\caption{Raw data used for the model in equation xxx and Figure xxx of the main article. Color codes the number of RAC exceedances (on a log-scale). Grey points denote sites without any exceedance.}
	\label{fig:ezgagrirac}
\end{figure}
\todo{update refs}
